Estimating duration in the presence of misclassification

### 2024-03-07 biology-data.tex
Overleaf comments

The Ct value decreases as the amount of DNA present in the original sample increases~\autocite{hanRTPCR}.
If there is more RNA present, more DNA is created, and hence fewer cycles are required to reach detection.
As the number of DNA copies doubles each cycle, the Ct value decreases increases linearly on the log-scale each cycle (see \cref{biology-data:fig:ct-calibration}).
A \emph{standard curve} translates the Ct value to viral load for a given test.
They are produced by measuring the Ct value using reference samples with known viral load.
As seen in \cref{biology-data:fig:ct-calibration}, they vary between studies.
\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{biology-data/ct-calibration}
    \caption[Example standard Ct curves]{%
        Example standard curves, which relate the logarithm of the viral copy number (the number of copies of the target DNA or RNA in the sample) to the measured Ct value.
        Reproduced from \textcite{hanRTPCR} with permission.
    }
    \label{biology-data:fig:ct-calibration}
\end{figure}

>>>>

The Ct value decreases as the amount of DNA present in the original sample increases~\autocite{hanRTPCR}.
If there is more RNA present, more DNA is created, and hence fewer cycles are required to reach detection.
As the number of DNA copies doubles each cycle, the Ct value decreases increases linearly on the log-scale each cycle (see \cref{biology-data:fig:ct-calibration}).
A \emph{standard curve} translates the Ct value to viral load for a given test.
They are produced by measuring the Ct value using reference samples with known viral load, and, as seen in \cref{biology-data:fig:ct-calibration}, they vary between studies.
\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{biology-data/ct-calibration}
    \caption[Example standard Ct curves]{%
        Example standard curves, which relate the logarithm of the viral copy number (the number of copies of the target DNA or RNA in the sample) to the measured Ct value.
        Reproduced from \textcite{hanRTPCR} with permission.
    }
    \label{biology-data:fig:ct-calibration}
\end{figure}

### 2024-03-07 introduction.tex
Overleaf comments

COVID-19 (coronavirus disease 2019), the disease caused by the virus SARS-CoV-2 (severe acute respiratory syndrome coronavirus 2), is estimated to have killed approximately 15 million in the acutest phase (to the end of 2021) and 30 million in total~\autocite{whoCOVIDExcess,economistCOVIDExcess}.
While would be reassuring if this was a rare or extreme event, but history suggests otherwise (see \cref{intro:fig:pandemic-timeline}).
\begin{figure}
    \includegraphics[width=\textwidth]{introduction/pandemic-timeline}
    \caption[Timeline of pandemics.]{%
        Timeline of pandemics.
        See \url{https://ourworldindata.org/historical-pandemics} for details and sources.
        Licensed under \href{https://creativecommons.org/licenses/by/4.0/}{CC-BY} by Saloni Dattani, Klara Auerbach, Marwa Boukarim, and Max Roser.
    }
    \label{intro:fig:pandemic-timeline}
\end{figure}

@@

All pandemics start as an \emph{outbreak}, a sudden and unexpected rise in the number of people with a disease in a local area; if this spreads over a larger geographic area (\eg a country) it becomes an \emph{epidemic}; if it spreads globally then it is a \emph{pandemic}~\autocite{grennanPandemic}.
The precise meaning of these terms is contested, determined by usage rather than formal definitions~\autocite{morensPandemic,doshiElusive}.
I will refer to the SARS-CoV-2 epidemic within England (or the UK), and the global impact as the pandemic.

In this context, \emph{surveillance} is the ability to detect and monitor the spread of an outbreak, epidemic, or pandemic.
Surveillance is important at every stage: early on, for detecting a new outbreak quickly and assessing its likely impact and later for monitoring the ongoing burden of the disease and the effectiveness of interventions~\autocite{whoFluSurveillance}.
In all cases, surveillance is required to inform public health policy.

@@

Various forms of \emph{reproduction number} were prominent throughout the COVID-19 pandemic~\autocite{pellisEstimation}.
Broadly, a reproduction number describes the number of secondary infections generated by a single infection.
Here, following \textcite{pellisEstimation} (who themselves are building on over a century of infectious disease dynamics literature), I define three that are broadly useful.

@@

The \emph{effective reproduction number}, $\Re$, is the mean number of secondary infections generated by a single infection at time $t$.
In particular, it uses the behaviour and immunity status at time $t$.
With constant behaviour, $\Re$ declines smoothly as individuals in the population become immune due to infection.
$\Re$ is useful because it describes the current dynamics of the epidemic: $\Re > 1$ means the epidemic is growing, $\Re < 1$ means it is shrinking, and $\Re = 1$ means it is stable.
In the simplest epidemic models (see \cref{E-SEIR:sec:SIR}), $\Re$ is proportional to the product of the rate of contacts and the proportion of the population that is susceptible.
Therefore, it informs the strength of intervention (either reducing contacts or amount of immunisation) required to control the epidemic.

The \emph{control reproduction number}, $\Rc$, is the mean number of secondary infections generated by a single infection with the behaviour at time $t$ but in a fully susceptible population.
$\Rc$ is also known as the \emph{reproduction number excluding immunity}.
It describes the effect on the epidemic of interventions (except vaccination) that have been put into place.
$\Rc$ is useful when comparing the effects of interventions at different points in time, when the immunity status of the population has changed.
For example, when comparing the English lockdown in March 2020 and Janurary 2021, there are two effects in play: that there are more individuals with immunity in January 2021, and that the lockdown had different measures in place.
$\Rc$ focuses on the latter effect.

@@

To translate incidence into the burden of the disease, a measure of severity is required.
For example, the IFR (infection fatality rate), the proportion of infected individuals who die.
Estimation of severity is beyond the scope of this thesis.
% Severity does not feature in this thesis, but it is important to note that it is a key component of understanding the burden of a pandemic.\todo{maybe reword sentence}

@@

Community testing is broadly available, but does not test a random sample of the population.
The number of tests conducted varies over time and space, for a variety of known and unknown reasons.
These include: policy on who is eligible for testing, the availability of tests, and the willingness of individuals to be tested.
Therefore, fluctuations in the number of positive tests may not reflect changes in the true incidence of the disease.
For large parts of the pandemic, community testing was widely available in the UK meaning that this is a large dataset, allowing detailed analyses to be performed, including stratification by small geographic areas or age groups.

Seroprevalence testing is a marker of recent prior infection.
Prior infection is more common than current infection, hence seroprevalence estimation requires a smaller sample size than current prevalence.
However, there is a delay between infection and the development of antibodies, hence seroprevalence is not a real-time measure of the burden of the disease.

Severe events are generally a reliable measure of the burden of the disease.
The criteria for being a severe event are well-defined and generally well-recorded in the UK context.
Therefore, the data is subject to fewer biases than community testing.
However, SARS-CoV-2 rarely leads to severe events in younger individuals, meaning severe events only capture transmission to a subset of the population.

@@

The use of prevalence surveys for estimating incidence and transmission is the focus of this thesis.
In particular, I focus on the CIS (Coronavirus (COVID-19) Infection Survey), run by the Office for National Statistics (ONS)~\autocite{CIS,cisMethodsONS}.
The CIS was a longitudinal study, testing a large number of indidividuals weekly or monthly for SARS-CoV-2; it is described in detail in \cref{E-intro:sec:cis}.
Early in the pandemic, a methodology was developed for using CIS data to estimate prevalence (see \cref{E-intro:sec:cis}).

A second prevalence survey ran in the UK, known as REACT (REal-time Assessment of Community Transmission)~\autocite{rileyResurgence,rileyREACT}.
REACT was a series of cross-sectional surveys, run across 19 rounds between May 2020 and March 2022.
Each round consisted of at least 100,000 randomly-sampled individuals from the UK contributing swabs for RT-PCR testing and questionnaire data~\autocite{elliottTwin}.

A major advantage of the CIS over REACT is that the CIS has daily testing of individuals, while REACT has no test results between rounds.
This means that REACT cannot be used to estimate transmission or incidence between rounds, instead relying on model assumptions at these times.
REACT missed the peak in late December 2020, which complicated analysis of the survey data at that time~\autocite{rileyREACTround8}.

The primary aim of this thesis is to estimate incidence and transmission from the CIS data.

@@

As noted above, incidence and transmission are extremely important quantities for informing the response to a pandemic but cannot be observed directly.
Therefore, they must be estimated using some form of modelling.

@@

\emph{Backcalculation} is a statistical procedure for estimating incidence based on a measurement of prevalence.
It was proposed by \textcite{brookmeyerMethod} in the context of HIV/AIDS.
In this context, backcalculation estimates the incidence of HIV infection from data on the number of AIDS diagnoses, and the time between infection and AIDS diagnosis~\autocite{brookmeyerBackcalculation,brookmeyerMeasuring}.
Two processes needed to be considered in the context of HIV/AIDS: one generating infections and one relating infections to prevalence.
% AIDS is incurable, therefore, AIDS prevalence is equivalent to cumulative incidence.

The statistical models used to describe HIV transmission are applicable to SARS-CoV-2.
This infection process can be described as a Poisson process (see \cref{E-inc-prev:sec:infection-process}).

Unlike HIV/AIDS, recovery from SARS-CoV-2 happens in the vast majority of infections.
Therefore, recoveries need to be considered in the backcalculation.
This introduces a dependency structure in the prevalence data (see \cref{E-inc-prev:sec:prevalence-process}).
The development of this framework makes clear that the duration of positivity (how long an individual tests positive for SARS-CoV-2) is the key quantity relating prevalence to incidence.
Therefore, estimation of duration is central to this thesis (see \cref{intro:sec:duration}).

Furthermore, prevalence surveys are noisy because they are based on population sampling.
This noise needs to be considered in the application of backcalculation (see \cref{E-inc-prev:sec:observation-process}).

@@

Mechanistic models make assumptions on the biological and epidemiological mechanisms driving transmission~\autocite{lesslerMechanistic}.
Therefore, the model's state and parameters have a biological and/or epidemiological interpretation.
% Of particular interest is that the basic reproduction number, $\R$, and the effective reproduction number, $R_e(t)$, which can both be calculated as functions of the model parameters; using \cref{E-backcalc}'s approach, additional modelling would be required to estimate these numbers.
In particular, this allows changes in transmission to be decomposed into the contributing factors, \eg the number of contacts per day or the probability of transmission upon contact (see \cref{E-SEIR:sec:mechanistic-models}).
The ability to estimate the contribution of each of these factors can inform public health policy.
Furthermore, mechanistic models can be used for scenario modelling, which simulates the effect of interventions.
A scenario is the effect of a proposed intervention (or lack thereof), simulated by modifying the relevant parameters.
% The mechanistic model can then be simulated forward in time to understand the intervention's effects.
% All of these properties make mechanistic models a useful tool for understanding and controlling infectious diseases.
% However, they make stronger assumptions about the disease and population behaviour than statistical models.

@@

Many mechanistic modelling studies were conducted during the pandemic, with at least 100 articles published in the first six months of 2020 alone~\autocite{shankarSystematic}.
In the UK, SPI-M-O (Scientific Pandemic Influenza Group on Modelling, Operational sub-group) was a sub-committee of SAGE (the Scientific Advisory Group for Emergencies), generating evidence based on modelling for policy development~\autocite{medleySPIM,govSPIMO}.
For example, the stages of lifting the lockdown in 2021 each had a set of modelling studies informing the government's decision~\autocite{sageEvidence}.

Each week, the members of SPI-M would produce a set of estimates of the current state of the epidemic (\eg the value of $\Re$), known as \emph{nowcasting}.
The nowcasting estimates were based on mechanistic models.
These estimates, once any estimates unduly affected by issues (\eg data quality issues), were statistically combined into a consensus value~\autocite{parkCombining}.
The consensus value was published on \url{gov.uk} and informed the government's response to the epidemic~\autocite{govRnumber}.
The academic groups contributing estimates used a variety of models, mostly based on either \emph{renewal equations} or \emph{compartmental models}.
The methods were reviewed by \textcite{royalSocietyRnumber}; see appendix 2 of that review for details of the specific implementations.

The renewal equation~\autocite{fraserRenewal} for an infectious disease, in the most general form, is:
\begin{align}
    I(t) &= \int_{0}^{t} \beta(t,\tau) I(t-\tau) d \tau
\end{align}
where $I(t)$ is the mean incidence at time $t$ and $\beta(t, \tau)$ is the instantaneous probability of an individual infected $\tau$ time units ago infects an individual at time $t$.
A scaled measure of $I(t)$ (\eg hospital admissions or community test results) allows estimating $\beta$.
The most common implementations are the R packages EpiNow2~\autocite{EpiNow2} and EpiEstim~\autocite{EpiEstim}.
Renewal equations do not feature in this thesis, for further information, including practical considerations, see \textcite{nashEstimating,gosticPractical}.

Compartmental models split the population into compartments, each representing a different state of the disease.
The \emph{dynamics} of the epidemic, how it changes over time, are described by the transitions between compartments.
Mathematically, the dynamics are described by ODEs (ordinary differential equations) or SDEs (stochastic differential equations).
Unlike renewal equations, there are not widely-used implementations of compartmental models.
Most of the implementations used by SPI-M-O members were in the SEIR (susceptible, exposed, infectious, recovered) framework.
The SEIR framework, has four components (susceptible, exposed, infectious, recovered), with the population divided into these components.
The SEIR framework is described in detail in \cref{E-SEIR:sec:mechanistic-models}.
Compartmental models can be fit to almost any data source, and commonly incorporate multiple data sources, known as \emph{data synthesis}, a technique reviewed by \textcite{birrellEvidence}.
Most compartmental models used by SPI-M-O were age-stratified, as differences between age groups are important for both the transmission and severity of SARS-CoV-2 (see \cref{E-SEIR:sec:structured-populations}).

Several compartmental models incorporate data from SARS-CoV-2 prevalence surveys, alongside other data sources~\autocite{daviesAssociation,ironsEstimating,knockKey,nicholsonImproving,pooleyEstimation,birrellRTM2}.
These models incorporate prevalence survey data by adding additional compartments representing RT-PCR individuals, who may not be infectious (this is common, see \cref{E-biology-data}); except \textcite{nicholsonImproving} who have a more sophisticated approach (below) and \textcite{ironsEstimating}, who assume RT-PCR positivity coincides with infectiousness.
The length of stay in these compartments (the duration of time positive) is assumed to follow a gamma distribution, with the parameters estimated from external data or having at most one free parameter.
The proportion of the population in these compartments are then fit to the prevalence survey data using a binomial likelihood where possible, although some approaches take a Gaussian approximation.
These approaches are similar to how these models are fit to other data sources, such as hospital admissions or deaths.

\Textcite{nicholsonImproving} have the most sophisticated use of prevalence survey data.
They allow an arbitrary distribution of the duration of positivity, based on the estimates of \textcite{hellewellPCRSensitivity} (see \cref{intro:sec:duration-SARS-CoV-2} for discussion of the \textcite{hellewellPCRSensitivity} study).
Furthermore, they incorporate the sensitivity and specificity of the RT-PCR test into their model.
% However, they do not incorporate age-stratification into their transmission model, and primarily aim to estimate the bias of community testing in the UK rather than incidence and/or transmission.

In \cref{E-SEIR}, I describe a compartmental model for SARS-CoV-2 transmission, which I fit to CIS data.
I allow a more flexible distribution of the duration of positivity than most of the papers discussed above, incorporating the skew of the distribution.
This distribution is based on data from a representative sample of the UK population, rather than convenience samples (see \cref{intro:sec:duration-SARS-CoV-2}).
Unlike \cref{nicholsonImproving}, I include age-stratification in the transmission model, allowing estimation of how transmission varies by age.
I use only data from the CIS, and show this is sufficient for estimating incidence and transmission.
Finally, I fit directly to counts from the CIS, rather than an approximation of this data, which has not previously be done.

@@

\todo[inline]{Introduce estimating duration}

@@

Modelling the duration directly, as I do in \cref{E-perf-test,E-imperf-test} considers only the binary result at each test (positive or negative); in this thesis a \emph{survival analysis} framework is adopted.
This approach generally requires fewer parameters, and no random effect, reducing the computational burden.
It is also easier to specify non-parameteric forms of the distribution, which is important for SARS-CoV-2 as the tail behaviour may be important.
A more detailed review of the relevant literature is in the relevant chapters.
% Survival analysis is the area of statistics concerned with estimating the distribution of the times between two events.
Other related approaches, such as a multi-state model~\autocite{jacksonMSM}\todo{find best reference for MSMs}, could also be considered (multi-state models can be viewed as a generalisation of survival analysis).

@@

% In addition to population-level surveillance, cohort studies help understand other aspects of the pathogen~\autocite{royalSocietyRnumber}.
% For example, contact tracing data can inform how likely it is for transmission to occur in different settings and/or between different groups of people.
The first study I use, ATACCC (Assessment of Transmission and Contagiousness of COVID-19 in Contacts)~\autocite{singanayagamCommunity,hakkiOnset}, performed daily testing of a small number of individuals with short follow-up.
The density of this sampling allows for reliable estimates of the viral load dynamics and other quantities of interest, most importantly the duration of detectability, but the results are limited because the small sample size and short follow-up means that the tails of these distributions cannot be estimated.
I describe ATACCC in detail in \cref{E-biology-data:sec:ataccc} and use it to estimate duration in \cref{E-ATACCC}.

The second study is the CIS, which I previously mentioned.
In addition to estimating prevalence, the CIS followed the same individuals over a long period of time (over a year in many cases), collecting longitudinal data.
The large sample size and long follow-up mean that the tails of the distribution of the duration of positivity can be estimated.
However, the lack of frequent testing (at most weekly) means that the viral load dynamics cannot be estimated.
In particular, the vast majority of detected infections had only one or two positive tests, and hence the viral load trajectory of these infections cannot be estimated.
I describe the CIS in detail in \cref{E-intro:sec:cis} and use it to estimate duration in \cref{E-perf-test,E-imperf-test}.

@@

For the purposes of this thesis, estimates applicable to the general population are required.
Therefore, I estimate the duration of positivity from the ATACCC and CIS studies, as described above.

@@

All inference in this thesis is Bayesian.
Bayesian inference has several relevant advantages.
Prior information allows identifiability in situations where this would not otherwise be possible, allows for combining knowledge from multiple sources, and many previous studies I build used Bayesian inference.

@@

In order to achieve these aims, I use the SARS-CoV-2 epidemic in England from Mon 31st Aug 2020 until Sun 25th Jan 2021 as a case study.
This case study has several features that make it a useful case study for the aims of this thesis.
First, the CIS was of sufficient size and incidence sufficiently high that there is enough data to perform the analyses.
Second, incidence begins rising over this period, which is the period of most interest.
In particular, an epidemic which will cause issues starts by increasing, therefore if the methods are to be applied early in a future epidemic or pandemic, they will start with an increasing period.
\todo{Rephrase this sentence}
The CIS began in April 2020, when incidence was falling due to the first lockdown in the UK.
Incidence remained low and fairly flat in England over summer 2020, until it began rising in mid or late August.
\todo{cite the phases of the UK epidemic}
Finally, the period ends before the roll-out of the vaccine, which likely affects many important aspects of the analysis.
In particular, the duration of positivity and the susceptibility of the population.
Including both of these elements would require more complex modelling but include idiosyncratic features of the SARS-CoV-2 pandemic that are probably less generalisable to future diseases.

@@

First, the framework to simulation the CIS (used in \cref{E-imperf-test:sec:simulate,E-perf-test:sec:simulation-study}).
The package enables customisation of various features of the study, for example the testing regime of the participants, the sensitivity of the tests, and the duration of positivity.
This package would be particularly useful in answering questions regarding the design of future prevalence surveys, for example, how frequently should individuals be tested, or how many individuals should be tested.

@@

First, in \cref{E-inc-prev:sec:observation-process}, I provide a theoretic justification for the use of deterministic backcalculation in this setting.
A deterministic backcalculation approach was used by \textcite{abbottCISincidence} and I will use it in \cref{E-backcalc}.
However, it assumes that the prevalence at any time is a random variable, instead modelling it as a deterministic function of the incidence.
\todo{Rephrase this sentence}
I show that, for all that is required for this assumption to hold is that the proportion of the population sampled is small.
In particular, I show that there is no requirement for the prevalence or incidence to be large.

>>>>

The coronavirus disease 2019 (COVID-19) caused by the Ssevere acute respiratory syndrome coronavirus 2 (SARS-CoV-2), is estimated to have killed approximately 15 million in the acutest phase (to the end of 2021) and 30 million globally~\autocite{whoCOVIDExcess,economistCOVIDExcess}.
While it is tempting to consider this a rare or extreme event, history suggests otherwise (see \cref{intro:fig:pandemic-timeline}).
\begin{figure}
    \includegraphics[width=\textwidth]{introduction/pandemic-timeline}
    \caption[Timeline of pandemics.]{%
        Timeline of pandemics.
        See \url{https://ourworldindata.org/historical-pandemics} for details and sources.
        Licensed under \href{https://creativecommons.org/licenses/by/4.0/}{CC-BY} by Saloni Dattani, Klara Auerbach, Marwa Boukarim, and Max Roser.
    }
    \label{intro:fig:pandemic-timeline}
\end{figure}

@@

\todo[inline]{Maybe talk about pandemic v epidemic v outbreak terminology?}
\todo[inline]{Need to motivate the importance of surveillance}

@@

Various forms of \emph{reproduction number} were prominent throughout the COVID-19 pandemic~\autocite{pellisEstimation}.
Broadly, a reproduction number describes the number of secondary infections generated by a single infection.
Here, following \textcite{pellisEstimation},  building on over a century of infectious disease dynamics literature, I define three that are broadly useful.

@@

The \emph{effective reproduction number}, $\Re$, is the mean number of secondary infections generated by a single infection at time $t$.
In particular, it uses the behaviour and immunity status at time $t$.
With constant behaviour, $\Re$ declines smoothly as individuals in the population become immune due to infection.
$\Re$ describes the current dynamics of the epidemic: $\Re > 1$ means the epidemic is growing, $\Re < 1$ means it is shrinking, and $\Re = 1$ means it is stable.
In the simplest epidemic models (see \cref{E-SEIR:sec:SIR}), $\Re$ is proportional to the product of the rate of contacts and the proportion of the population that is susceptible.
Therefore, it informs the strength of intervention (either reducing contacts or amount of immunisation) required to control the epidemic.

The \emph{control reproduction number}, $\Rc$, is the mean number of secondary infections generated by a single infection with the behaviour at time $t$ but in a fully susceptible population.
$\Rc$ is also known as the \emph{reproduction number excluding immunity}.
It describes the effect on the epidemic of interventions (except vaccination) that have been put into place.
$\Rc$ is useful when comparing the effects of interventions at different points in time, when the immunity status of the population has changed.
For example, when comparing the English lockdown in March 2020 and January 2021, there are two effects in play: that there are more individuals with immunity in January 2021, and that the lockdown had different measures in place.
$\Rc$ focuses on the latter effect.

@@

To translate incidence into the burden of the disease, a measure of severity is required.
For example, the infection fatality rate (IFR), the proportion of infected individuals who die.
Estimation of severity is beyond the scope of this thesis.
% Severity does not feature in this thesis, but it is important to note that it is a key component of understanding the burden of a pandemic.\todo{maybe reword sentence}

@@

Community testing is broadly available, but does not test a random sample of the population.
The number of tests conducted varies over time and space, for a variety of known and unknown reasons.
These include: policy on who is eligible for testing, the availability of tests, and the willingness of individuals to be tested.
Therefore, fluctuations in the number of positive tests may not reflect changes in the true incidence of the disease.
For large parts of the pandemic, community testing was widely available in the UK meaning resulting in a rich dataset, allowing detailed analyses to be performed, including stratification by small geographic areas or age groups.

Seroprevalence testing provides information on a marker of recent prior infection, by measuring ..????
Prior infection is more common than current infection, hence seroprevalence estimation requires a smaller sample size than current prevalence.
However, there is a delay between infection and the development of antibodies, hence seroprevalence is not a real-time measure of the burden of the disease.

Severe events are generally a reliable measure of the burden of the disease.
The criteria for being a severe event are well-defined and generally well-recorded in the UK context (reference?).
Therefore, the data is subject to fewer biases than community testing.
However, SARS-CoV-2 rarely leads to severe events in younger individuals, meaning severe events only capture sequelae of transmission in a subset of the population.

@@

The use of prevalence surveys for estimating incidence and transmission is the focus of this thesis.
In particular, I focus on the Coronavirus (COVID-19) Infection Survey (CIS ), run by the Office for National Statistics (ONS)~\autocite{CIS,cisMethodsONS}.
The CIS was a longitudinal study, testing a large number of indidividuals weekly or monthly for SARS-CoV-2; it is described in detail in \cref{E-intro:sec:cis}.
Early in the pandemic, a methodology was developed for using CIS data to estimate prevalence (see \cref{E-intro:sec:cis}).
The primary aim of this thesis is to estimate incidence and transmission from this data.

@@

As noted above, incidence and transmission are extremely important quantities for informing the response to a pandemic but cannot be observed directly.
Therefore, they must be estimated.

@@

\emph{Backcalculation} is a statistical procedure for estimating incidence based on a measurement of prevalence.
It was proposed by \textcite{brookmeyerMethod} in the context of HIV/AIDS.
In this context, backcalculation estimates the incidence of HIV infection from data on the number of AIDS diagnoses, and the time between infection and AIDS diagnosis~\autocite{brookmeyerBackcalculation,brookmeyerMeasuring}.
Two processes needed to be considered: one generating infections and one relating infections to prevalence.
% AIDS is incurable, therefore, AIDS prevalence is equivalent to cumulative incidence.

The statistical framework developed for HIV is applicable to SARS-CoV-2.
(see \cref{E-inc-prev:sec:infection-process}).

However, Unlike HIV/AIDS, recovery from SARS-CoV-2 infection is possible, so recoveries need to be considered in the backcalculation.
This introduces a dependency structure in the prevalence data (see \cref{E-inc-prev:sec:prevalence-process}).
The development of this framework makes clear that the duration of positivity (how long an individual tests positive for SARS-CoV-2) is the key quantity relating prevalence to incidence.
Therefore, estimation of duration is central to this thesis (see \cref{intro:sec:duration}).

Furthermore, prevalence surveys are noisy because they are based on population sampling.
This noise needs to be considered in the development of the backcalculation approach (see \cref{E-inc-prev:sec:observation-process}).

@@

Mechanistic models make assumptions on the biological and epidemiological mechanisms driving transmission~\autocite{lesslerMechanistic}.
Therefore, the model's state and parameters have a biological and/or epidemiological interpretation.
% Of particular interest is that the basic reproduction number, $\R$, and the effective reproduction number, $R_e(t)$, which can both be calculated as functions of the model parameters; using \cref{E-backcalc}'s approach, additional modelling would be required to estimate these numbers.
In particular, this allows changes in transmission to be decomposed into the contributing factors, \eg the number of contacts per day or the probability of transmission upon contact (see \cref{E-SEIR:sec:mechanistic-models}).
The ability to estimate the contribution of each of these factors can inform public health policy.
Furthermore, mechanistic models can be used for scenario-based modelling, which simulates the effect of interventions.
A scenario here is the effect of a proposed intervention (or lack thereof), introduced into the model by modifying the relevant parameters.
% The mechanistic model can then be simulated forward in time to understand the intervention's effects.
% All of these properties make mechanistic models a useful tool for understanding and controlling infectious diseases.
% However, they make stronger assumptions about the disease and population behaviour than statistical models.

@@

Many mechanistic modelling studies were conducted during the pandemic, with at least 100 articles published in the first six months of 2020 alone~\autocite{shankarSystematic}.
In the UK, the most influential were the outputs from SPI-M\todo{cite SPI-M and its outputs - Medley paper?}.
For example, the stages of lifting the lockdown in 2021 each had a set of modelling studies informing the government's decision\todo{cite roadmap 2021 papers}.

Each week, the members of SPI-M would produce a set of estimates of the current state of the epidemic (\eg ther value of $\Re$), known as \emph{nowcasting}.
These estimates, once any estimates unduly affected by issues (\eg data quality issues), were statistically combined into a consensus value.\todo{cite SPI-M combination}
The consensus value was published on \url{gov.uk} and informed the government's response to the epidemic.\todo{cite gov.uk R number}

The academic groups contributing estimates used a wide variety of methodologies

\todo[inline]{Finish lit review of SPI-M methods}

\begin{itemize}
    \item SPI-M
    \item Literature on mechanistic models using REACT and CIS I have previously circulated.
\end{itemize}

@@

@@

Modelling the duration directly, as I do in \cref{E-perf-test,E-imperf-test} considers only the binary result at each test (positive or negative); in this thesis a \emph{survival analysis} framework is adopted.
This approach generally requires fewer parameters, and no random effect, reducing the computational burden.
It is also easier to specify non-parameteric forms of the distribution, which is important for SARS-CoV-2 as the tail behaviour may be important.
A more detailed review of the relevant literature is in the relevant chapters.
% Survival analysis is the area of statistics concerned with estimating the distribution of the times between two events.
Other related approaches, such as a multi-state model~\autocite{jacksonMSM}, could also be considered (multi-state models can be viewed as a generalisation of survival analysis).

@@

% In addition to population-level surveillance, cohort studies help understand other aspects of the pathogen~\autocite{royalSocietyRnumber}.
% For example, contact tracing data can inform how likely it is for transmission to occur in different settings and/or between different groups of people.
The first study I use, ATACCC (Assessment of Transmission and Contagiousness of COVID-19 in Contacts)~\autocite{singanayagamCommunity,hakkiOnset}, performed daily testing of a small number of individuals with short follow-up.
The density of this sampling allows for reliable estimates of the viral load dynamics and other quantities of interest, most importantly the duration of detectability, but the results are limited because the small sample size and short follow-up means that the tails of these distributions cannot be estimated.
I describe ATACCC in detail in \cref{E-biology-data:sec:ataccc} and analyse it in \cref{E-ATACCC}.

The second study is the CIS, which I previously mentioned.
In addition to estimating prevalence, the CIS followed the same individuals over a long period of time (over a year in many cases), collecting longitudinal data.
The large sample size and long follow-up mean that the tails of the distribution of the duration of positivity can be estimated.
However, the lack of frequent testing (at most weekly) means that the viral load dynamics cannot be estimated.
In particular, the vast majority of detected infections had only one or two positive tests, and hence the viral load trajectory of these infections cannot be estimated.
I describe the CIS in detail in \cref{E-intro:sec:cis} and analyse it in \cref{E-perf-test,E-imperf-test}.

@@

@@

All inference in this thesis is performed within the Bayesian framework.
Bayesian inference has several advantages relevant for this work.
Prior information allows identifiability in situations where this would not otherwise be possible, allows for combining knowledge from multiple sources, and much prior work I build on has been performed within the Bayesian framework.

@@

In order to achieve these aims, I use the SARS-CoV-2 epidemic in England from Mon 31st Aug 2020 until Sun 25th Jan 2021 as a case study.
This case study has several features that make it a useful case study for the aims of this thesis.
First, the CIS was of sufficient size and incidence sufficiently high that there is enough data to perform the analyses.
Second, incidence begins rising over this period, which is the period of most interest.
In particular, an epidemic which will cause issues starts by increasing, therefore if the methods are to be applied early in a future epidemic or pandemic, they will start with an increasing period.
The CIS began in Aril 2020, when incidence was falling due to the first lockdown in the UK.
Incidence remained low and fairly flat in England over summer 2020, until it began rising in mid or late August.
\todo{cite the phases of the UK epidemic}
Finally, the period ends before the roll-out of the vaccine, which likely affects many important aspects of the analysis.
In particular, the duration of positivity and the susceptibility of the population.
Including both of these elements would require more complex modelling but include idiosyncratic features of the SARS-CoV-2 pandemic that are probably less generalisable to future diseases.

@@

First, the framework to simulation the CIS (used in \cref{E-imperf-test:sec:simulate,E-perf-test:sec:simulation-study}) is structured as a R package.
The package enables customisation of various features of the study, for example the testing regime of the participants, the sensitivity of the tests, and the duration of positivity.
This package would be particularly useful in answering questions regarding the design of future prevalence surveys, for example, how frequently should individuals be tested, or how many individuals should be tested.

@@

First, in \cref{E-inc-prev:sec:observation-process}, I provide a theoretic justification for the use of deterministic backcalculation in this setting.
A deterministic backcalculation approach was used by \textcite{abbottCISincidence} and I will use it in \cref{E-backcalc}.
However, it assumes that the prevalence at any time is a random variable, instead modelling it as a deterministic function of the incidence.
I show that, for all that is required for this assumption to hold is that the proportion of the population sampled is small.
In particular, I show that there is no requirement for the prevalence or incidence to be large.
